|

### 2024-09-03 reproducibility/manuscript/manuscript.qmd
latex label bugfix

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{equation}
  \label{eq-rna}
\begin{aligned}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
\end{aligned}
\end{equation}
where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$. We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
\begin{equation}
    \label{eq-rna-cg}
\begin{aligned}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-du-cg} \\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-ds-cg} .
\end{aligned}
\end{equation}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u_{cg}(t), s_{cg}(t) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene.
We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression. We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown transcription state. We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$. These relative times are also used to parametrize the Bernoulli process for $k_{cg}$. 
Importantly, we recognize that the initial conditions are, in fact, unknown.

>>>>

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{equation}
  \label{eq-rna}
\begin{aligned}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
\end{aligned}
\end{equation}
where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$. We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
\begin{equation}
    \label{eq-rna-cg}
\begin{aligned}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} .
\end{aligned}
\end{equation}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u_{cg}(t), s_{cg}(t) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene.
We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression. We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown transcription state. We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$. These relative times are also used to parametrize the Bernoulli process for $k_{cg}$. 
Importantly, we recognize that the initial conditions are, in fact, unknown.
